---
title: "Data model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data model}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
# library(edd)
```

EDD

About EDD

EDD (Economic Data Dashboard) is an R package with an integrated Shiny
app. Much of the functionality of the R package is specifically designed
to deliver the Shiny app, but many of the additional included functions
will be useful to economists and analysts in their routine work, and
there are many others designed to be used outside of the Shiny app
itself in an interactive environment.

The EDD Shiny app is hosted at
<https://economic-analytics.shinyapps.io/edd-app>, and can be launched
from an interactive R session by running `edd::launch_edd()`.

# Data model

Built in R with front end UI in Shiny (Bootstrap), EDD is designed to
not be reliant on external databases for data, instead using in-memory
features of R.

Data is grouped into core datasets, most of which follow their
designation from the Office for National Statistics or their publisher.
These are named by their common abbreviations, where available, e.g.
DIOP for Index of Production, DRSI for Retail Sales Index, MGDP for
monthly GDP, etc. A full list of current and proposed datasets is
available in the data section.

All data is first transformed into single flat tables (data frames), in
normalised form (3NF, or Hadley Wickham's Tidy Data definition) to
provide native integration in many of R and the Tidyverse's core
packages.

## Dimensionality

Much economic data is highly dimensional, i.e. there are often many
variables describing geography, industry, business size, legal
structure, age, sex and many other characteristics. These are always
returned in the EDD data model as columns for each dimension, regardless
of the number of possible options for each dimension. This means that
for each row in a table, there is only one measured (numerical) value;
all other dimension data is returned as separate variables, including
date.

## Dates

Dates are always stored as Date objects (`dates$date`) and returned in ISO8601 format
(YYYY-MM-DD). To ensure that different frequencies of the same data are
identifiable (e.g. much data is available as monthly, quarterly and
annual series), there is a separate variable, `dates$freq`, that holds a
single character identifier ("m", "q", "a"), so that particular
frequencies can be distinguished. Datasets contain mixed frequency data
by design and are only separated into distinct frequency data frames
(actually `tsibble`s) at the point of any modelling or forecasting.
Dates are always stored as the first date in the period to which they
refer, e.g. "2020-01-01" represents either January 2020, Q1 2020 or the
whole of 2020 depending on the contents of `freq`, "2022-10-01"
represents either October 2022 or Q4 2022. There is an included
function, `date_to_text()`, which translates a date in EDD's two-column
format to a string, e.g.:

```
date_to_text(data.frame(date = "2022-10-01", freq = "m"))
#> "October 2022"

dates <- data.frame(date = c("2022-07-01", "2022-10-01"),
                    freq = rep("q", 2))

date_to_text(dates)
#> "Q3 2022" "Q4 2022"
```

Whilst maintaining data in a fully described "Tidy" format makes
integration into common R packages simple, it provides a challenge in
large datasets: high levels of duplication of dimensional data which
occupies significant amounts of memory where datasets are particularly
large and highly-dimensional. For this reason, before data flows to the
UI, all datasets are transformed to fifth normal form (5NF) to remove
all data redundancy and save memory.

Example

| Date       | Geography Code | Geography Name | Industry Code | Industry Name | Variable      | Value |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| 2022-04-01 | E08000003      | Manchester     | C             | Construction  | GVA CVM Index | 101.5 |

For a large time series, with all local authorities (>300), SIC2007
Sectors (20) and a number of variables, the data redundancy becomes
large. Data is transformed into 5NF to remove this:

```
data
date | geography.code | industry.code | variable | value

geography
geography.code | geography.name

industry
industry.code | industry.name

variable
variable.code | variable.name | variable.unit
```

As well as reducing both storage space and memory usage, it also allows
for lazy loading of the primary `data` table, as UI select boxes are
populated from the dimension lookup tables which store only distinct
values, reducing data transfer to the user.

As the user selects their dimension options, then the data table is
filtered through a series of joins to each dimension table. This means
the final data frame is built for UI analysis in 3NF ("tidy") contains
only those rows requested by the user and minimises redundancy until the
last moment.

## Nested data frames

R supports the use of nested data frames within data frames, i.e. a
column of a data frame can be another data frame whose number of rows
does not need to equal that of the parent (a key constraint on data
frames). EDD utilises this to group together multiple columns that
relate to the same variable, ensuring that all relevant information for
any particular dimension is not separated from others. For example, in
our initial 3NF data frame from above, the data is actually stored like
this:

```
dates       | geography   | industry    | variable           | value
date | freq | code | name | code | name | code | name | unit | 
```
Notice that value, the only numeric-type variable in any EDD data frame,
is the only "true" single variable column: the rest are all columns of
data frames that themselves form columns in the "master" data frame. In
R:
```
master <- data.frame(dates = data.frame(date, freq),
                     geography = data.frame(code, name),
                     industry = data.frame(code, name),
                     variable = data.frame(code, name, unit),
                     value
                     )
```
Whilst R natively supports the above ideas, they are more easily
expressed through the Tidyverse's `tibble`, which is the implementation
in EDD.

The advantage of this format is that all characteristics of all
dimensions are easily passed around between functions by simply
referring to the dimension, e.g. `master$geography` or
`master[["geography"]]` and all columns, regardless of their number, are
always passed.

Clearly with a large number of datasets, the number of subsidiary
dimension lookup tables grows rapidly and may cause confusion. To
address this, dimension lookup tables are connected to their parent data
table by their inclusion in the same list.

```
dataset <- list(
  data = tibble(dates = tibble(date, freq),
                industry,
                geography,
                variable,
                value),
  dimensions = list(
    geography = tibble(code, name),
    industry  = tibble(code, name),
    variable  = tibble(code, name, unit)
    )
)
```

This standardised data architecture underpins the EDD model. It is not
(yet) enforced through any OOP tooling, but effective constructor and
verification functions (`as_eddobj()`, `is_eddobj()`) will be available, as
well as functions to return this model to 3NF (`eddobj_to_edd_df()`) and
vice versa (`df_to_edd_df()`, `edd_df_to_edd_obj()`).

## Object verification

Verification of the variety of different objects that EDD uses is
undertaken through a set of `is_*()` functions that return a logical
vector of length 1. `is_df()`, `is_edd_df()` and `is_edd_obj()` are the
primary functions. Each of these objects is described in more detail
here.

Class `df` identifies a `data.frame` which has correctly named columns
and can be transformed into other objects for use inside EDD. The rules
that are tested to verify a class of `df` are: - existence of a column
`date` in one of the below formats: - `YYYY` -
`YYYY [month name (full or abbreviated)]` or
`[month name (full or abbreviated)] YYYY` - `YYYY Qn` or `Qn YYYY` If
none of these formats exist, EDD will prompt you for a format mask which
you can also supply in the `date_parse_dates()` function's `format`
argument. - existence of a column `value` which contains only numeric
values (or values coercible to numeric) - existence of a column
`variable` which contains only character values describing the value

This is the minimum acceptable `data.frame` that will verified with
`is_df()`.

Other columns can be added so long as they follow the EDD naming scheme:

Object construction

Any flat CSV or Excel file can be passed to the constructor function as
long as a number of conditions are met:

Compulsory fields:

Any CSV to be imported must contain, at a minimum, the following fields:
\* date \* variable \* value

This is the smallest amount of information that EDD requires to import
your data. Case does not matter as all field names are converted to
lower case at import.

If the data contains any dimensions, these should be in their own
separate columns, e.g.

`"date", "geography", "industry", "value"`

If any dimension has more than one field associated with it, e.g.
geography code, geography name, then these fields should be named in
this way: `"geography_code", "geography_name"`
